The reference model approach in feature selection problems \vspace{.1in}

### 2019-07-01 paper/ref_approach.tex
updated bodyfat real world example structure

In statistical applications, one of the main steps of the modelling workflow is covariate or feature selection, which is a special case of model reduction. The necessity of carrying out the selection arises from different problematics. Sometimes the leading objective of the analysis is to predict future observations, thus the main focus in building the model is on the predictive performance. Yet a high dimensional parameter space often leads to dangers as overfitting or other technical problems, as expensive costs to collect all the features for future observations or computational and memory burdens (e.g. when we are operating on limited resources), and thus a sparser representation of the model would be preferred. In addition to that, simpler models bring also advantages in terms of interpretability. Some other times, variable selection can be also seen as an inferential tool: the interest is on knowing which variables are meaningful, or relevant, in relation to the target variable of interest. This typically happens in a \textit{small n large p} scenario, as in many microarray data analysis in medicine or biology. A very large number of genes is usually analysed in two groups of individuals labeled as \textit{positive} or \textit{negative} to some disease; the goal is to spot the subset of explanatory genes in order to perform further experiments \cite[see examples in][]{efron2012large}.

@@

Figure \ref{fig:k} shows the posterior means for the shrinkage factors of each parameter using the regularised horseshoe prior, averaged after 100 data simulations. Relevant and non-relevant features are respectively highlighted in red and blue. The ideal shrinkage would have $\kappa=0$ when the parameter is a signal, whereas $\kappa=1$ when it is noise. Respectively on the y-axis and on the x-axis are depicted the shrinkage factors using the reference model approach or not. If the reference model did not make any difference, we would expect the points to lie on the diagonal. Non-relevant features happen to lie on average on the diagonal for all the simulated scenario, meaning not evident benefits coming from the reference model. Although when considering the shrinkage amount for the relevant ones, we note that the points always lie under the diagonal, meaning that the reference model is able to shrink less the true signals. Such benefit is proportional to the difficulty of the problem, that is it increases with lower correlation and number of observations. Note that the perfect distinction of the two clusters of features (relevant and non-relevant) observable on both the marginal distributions of the shrinkage factors is due to the averaging over the 100 data simulations. In a single data realisation there would be more noise and the clusters would be closer and more overlapping.
The results for the Dirichlet-Laplace prior are very similar and they are reported in the Appendix \ref{ap:shrinkage_signal}.

@@

We conclude our experiments using the real world dataset body fat, already described in the comparison between the projection predictive approach and the stepwise regression in Section \ref{projection}. Consistently with the previous comparisons, we design the variable selection by normal means problem. We add to the original data noisy uncorrelated covariates up to a total of 1000 features. These covariates are normally distributed and scaled as the original ones. We compute correlations between each variable and the target variable, that is the amount of fat, and transform them by Fisher transformation. The original assumption, in order that \eqref{eq:fisher_transformation} holds, is that the variables are jointly normally distributed. In our experience the normal approximation in \eqref{eq:fisher_transformation} is still reasonable, but after rescaling by $\sqrt{n-3}$ we do not fix the variance to be one, instead we estimate it from the data. The methods we compare are those used in Section \ref{complete-selection}: the control of the local false discovery rate (loc.fdr), the empirical Bayes median (EB.med) and the selection by posterior credible intervals at level 0.9 (ci.90). The maximum likelihood estimator is used to estimate the variance of the null hypothesis in the local false discovery rate \cite[see][Chap. 6]{efron2012large}, whereas the median absolute deviation from zero for the empirical Bayes median method. The selection via posterior credible intervals is done using a regularised horseshoe prior on the parameters for the means, while a standard log-normal  prior for the variance. In order to vary the difficulty of the selection, we bootstrap subsamples of different sizes, going from $n=50$ up to $n=251$. Each time results are averaged over 100 bootstrap subsamples.

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_sensitivity_vs_fdr.pdf}
  \caption{Sensitivity against false discovery rate after 100 data simulations.\\}
  \label{fig:bodyfat_sensitivity_vs_fdr}
\end{figure}

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_stability.pdf}
  \caption{Stability estimates with 0.95 confidence intervals after 100 data simulations.\\}
  \label{fig:bodyfat_stability}
\end{figure}

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_varyingP_sensitivity_vs_fdr.pdf}
  \caption{Sensitivity against false discovery rate after 100 data simulations.\\}
  \label{fig:bodyfat_varyingP_sensitivity_vs_fdr}
\end{figure}

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_varyingP_stability.pdf}
  \caption{Stability estimates with 0.95 confidence intervals after 100 data simulations.\\}
  \label{fig:bodyfat_varyingP_stability}
\end{figure}

As an additional experiment, we run the selection bootstrapping from the whole dataset, but varying the overall number of covariates from $p=100$ up to $p=5000$. Figures \ref{fig:bodyfat_varyingP_sensitivity_vs_fdr} and \ref{fig:bodyfat_varyingP_stability} respectively show the sensitivity against the false discovery rate and the stability results. As in the previous experiment, we note that the reference model has a larger impact as the complexity of the problem increases: previously as $n$ decreases, now as $p$ increases. Note the different scale of the sensitivity and the false discovery rate axes. We observe an odd behaviour of the selection via credible intervals (ci.90): when $p=100$ and $p=5000$ the results are strongly different from the other methods, both in terms of sensitivity/false discovery rate and stability. 
%Such behaviour can be in part explained by the connection between low stability and high uncertainty in the estimates.

@@

Finally, we repeate the selection of Section \ref{projection} implementing the stepwise backward selection by AIC as implemented in the \textit{step} function in R. We have not fixed any variable to be always included in the model and the overall number of covariates (original plus noisy) is 100. We compare results using or not the reference model on top of the procedure; here we have used again the reference model given by the first five supervised principal components. Figure \ref{fig:bodyfat_step_refvsdata} shows the number of noisy features included in the final model and the out-of-sample root mean square error (RMSE). Results are after 100 bootstrap simulations and the predictive performance is tested on the observations excluded at each bootstrap sample. We observe that the reference model reduces the number of noisy features included in the final model. This leads to less overfit to the data and, thus, improved out-of-sample predictive performance in terms of RMSE. Also in this case, we see that there is not a notable difference in terms of the stability of the two approaches, which in this case is qualitatively assessed through the variability of the histogram. We can not use the stability measure provided by \cite{paper:stability}, since it does not take in account possibly swaps in the set of selected features due to high correlation between the covariates, which in this case can occur. Our reference model approach applied to the stepwise backward selection achieves outstanding improvements considering its simplicity, yet it does not reach the goodness of more thoughtful implementations, as projpred (see results of Section \ref{projection}).

@@

This paper has discussed the importance of using a reference model when dealing with feature selection, or more in general model reduction. We want to highlight that the goal of this paper is not to provide a specific algorithm or procedure to carry out the selection, but rather to bring motivations and attentions to the general use of a reference model. 
\\
The reference model acts as an approximation of the data generation mechanism through its predictive distribution. Such approximation is generally less noisy than the sample estimation available from the observed data, leading to the main benefits of the reference model approach. In our comparisons, we have analysed the effect of a reference model in the form of a filter on the observed target values on top of different variable selection methods widely used. The overall benefits we have showed independently of the specific method applied, consisting in stability and goodness of the selection, motivate what we refer to as the reference model approach for the model reduction framework. Such approach translates into a family of different methods, some of which have been present in the literature for several years.
\\
In a real-world data application (body fat data), we have applied one of these methods, namely the projection predictive approach as indicated by \cite{paper:projpred}, in opposition to the selection via stepwise backward regression with the results provided by \cite{paper:bodyfat}. The selection via projection resulted to be more stable and led to a sparser representation. After adding additional noisy covariates, the selected submodel showed also higher predictive performance compared to the one selected by the stepwise backward regression, which included several non-relevant features.
\\
Whenever it is possible to come up with a reasonable reference model, our suggestion is to employ it when selecting models, because of its nature as baseline of the comparisons and its ability to improve the selection. Note that the main challenge in many real world application will consist in devising the reference model itself and assessing its predictive performance, for which we suggest the user to rely on robust modelling workflow indications \cite[see e.g.][]{gabry2019visualization}.
\\
All Bayesian models have been implemented using Stan \citep{paper:stan}, which uses HMC-NUTS sampler \citep{hoffman2014no} and allows full posterior Bayesian inference. The code to run all the experiments is available at \url{https://github.com/fpavone/ref-approach-paper}.

@@

\hypertarget{ap:shrinkage_signal}{%
\section*{Appendix A: Shrinkage factors with Dirichlet-Laplace prior}\label{ap:shrinkage_signal}}
Figure \ref{fig:k_DL} shows the posterior mean shrinkage factors using the Dirichlet-Laplace prior for the normal means problem in the simulated data study of Section \ref{comparison}. The result is qualitatively equal to the regularised horseshoe prior's one (see Figure \ref{fig:k}).

>>>>

In statistical applications, one of the main steps of the modelling workflow is covariate or feature selection, which is a special case of model reduction. The necessity of carrying out the selection arises from different problematics. Sometimes the leading objective of the analysis is to predict future observations, thus the main focus in building the model is on the predictive performance. Yet a high dimensional parameter space often leads to dangers as overfitting or other technical problems, as expensive costs to collect all the features for future observations or computational and memory burdens (e.g. when operating on limited resources), and thus a sparser representation of the model would be preferred. In addition to that, simpler models bring also advantages in terms of interpretability. Some other times, variable selection can be also seen as an inferential tool: the interest is on knowing which variables are meaningful, or relevant, in relation to the target variable of interest. This typically happens in a \textit{small n large p} scenario, as in many microarray data analysis in medicine or biology. A very large number of genes is usually analysed in two groups of individuals labeled as \textit{positive} or \textit{negative} to some disease; the goal is to spot the subset of explanatory genes in order to perform further experiments \cite[see examples in][]{efron2012large}.

@@

Figure \ref{fig:k} shows the posterior means for the shrinkage factors of each parameter using the regularised horseshoe prior, averaged after 100 data simulations. Relevant and non-relevant features are respectively highlighted in red and blue. The ideal shrinkage would have $\kappa=0$ when the parameter is a signal, whereas $\kappa=1$ when it is noise. Respectively on the y-axis and on the x-axis are depicted the shrinkage factors using the reference model approach or not. If the reference model did not make any difference, we would expect the points to lie on the diagonal. Non-relevant features happen to lie on average on the diagonal for all the simulated scenario, meaning not evident benefits coming from the reference model. Although when considering the shrinkage amount for the relevant ones, we note that the points always lie under the diagonal, meaning that the reference model is able to shrink less the true signals. Such benefit is proportional to the difficulty of the problem, that is it increases with lower correlation and number of observations. Note that the perfect distinction of the two clusters of features (relevant and non-relevant) observable on both the marginal distributions of the shrinkage factors is due to the averaging over the 100 data simulations. In a single data realisation there would be more noise and the clusters would be closer and overlapping.
The results for the Dirichlet-Laplace prior are very similar and they are reported in the Appendix \ref{ap:shrinkage_signal}.

@@

We conclude our experiments using the real world dataset body fat, already described in the comparison between the projection predictive approach and the stepwise regression in Section \ref{projection}. We carry out two different analyses: one using the normal means problem framework and the other one using the stepwise backward regression. In the former, we add to the original data noisy uncorrelated covariates up to a total of 1000 features. These covariates are normally distributed and scaled as the original ones. We compute correlations between each variable and the target variable, that is the amount of fat, and transform them by Fisher transformation. The original assumption, in order that \eqref{eq:fisher_transformation} holds, is that the variables are jointly normally distributed. In our experience the normal approximation in \eqref{eq:fisher_transformation} is still reasonable, but after rescaling by $\sqrt{n-3}$ we do not fix the variance to be one, instead we estimate it from the data. The methods we compare are those used in Section \ref{complete-selection}: the control of the local false discovery rate (loc.fdr), the empirical Bayes median (EB.med) and the selection by posterior credible intervals at level 0.9 (ci.90). The maximum likelihood estimator is used to estimate the variance of the null hypothesis in the local false discovery rate \cite[see][Chap. 6]{efron2012large}, whereas the median absolute deviation from zero for the empirical Bayes median method. The selection via posterior credible intervals is done using a regularised horseshoe prior on the parameters for the means, while a standard log-normal  prior for the variance. In order to vary the difficulty of the selection, we bootstrap subsamples of different sizes, going from $n=50$ up to $n=251$. Each time results are averaged over 100 bootstrap subsamples.

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_sensitivity_vs_fdr.pdf}
  \caption{Sensitivity against false discovery rate after 100 bootstrap samples.\\}
  \label{fig:bodyfat_sensitivity_vs_fdr}
\end{figure}

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_stability.pdf}
  \caption{Stability estimates with 0.95 confidence intervals after 100 bootstrap samples.\\}
  \label{fig:bodyfat_stability}
\end{figure}

@@

%\begin{figure}[tp]
%  \centering
%  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_varyingP_sensitivity_vs_fdr.pdf}
%  \caption{Sensitivity against false discovery rate after 100 data simulations.\\}
%  \label{fig:bodyfat_varyingP_sensitivity_vs_fdr}
%\end{figure}

%\begin{figure}[tp]
%  \centering
%  \includegraphics[width=0.98\textwidth]{graphics/bodyfat_varyingP_stability.pdf}
%  \caption{Stability estimates with 0.95 confidence intervals after 100 data simulations.\\}
%  \label{fig:bodyfat_varyingP_stability}
%\end{figure}

% As an additional experiment, we run the selection bootstrapping from the whole dataset, but varying the overall number of covariates from $p=100$ up to $p=5000$. Figures \ref{fig:bodyfat_varyingP_sensitivity_vs_fdr} and \ref{fig:bodyfat_varyingP_stability} respectively show the sensitivity against the false discovery rate and the stability results. As in the previous experiment, we note that the reference model has a larger impact as the complexity of the problem increases: previously as $n$ decreases, now as $p$ increases. Note the different scale of the sensitivity and the false discovery rate axes. We observe an odd behaviour of the selection via credible intervals (ci.90): when $p=100$ and $p=5000$ the results are strongly different from the other methods, both in terms of sensitivity/false discovery rate and stability. 
%Such behaviour can be in part explained by the connection between low stability and high uncertainty in the estimates.

@@

In the second analysis, we repeat the selection of Section \ref{projection} implementing the stepwise backward regression by AIC as implemented in the \textit{step} function in R. We have not fixed any variable to be always included in the model and the overall number of covariates (original plus noisy) is 100, as it was in the last part of Section \ref{projection}. We compare results using or not the reference model on top of the procedure; here we have used again the reference model given by the first five supervised principal components. Figure \ref{fig:bodyfat_step_refvsdata} shows the number of noisy features included in the final model and the out-of-sample root mean square error (RMSE). Results are after 100 bootstrap samples on the whole dataset, and the predictive performance is tested on the observations excluded at each bootstrap sample. We observe that the reference model reduces the number of noisy features included in the final model. This leads to less overfit to the data and, thus, improved out-of-sample predictive performance in terms of RMSE. Also in this case, we see that there is not a notable difference in terms of the stability of the two approaches, which in this case is qualitatively assessed through the variability of the histogram. We can not use the stability measure provided by \cite{paper:stability}, since it does not take in account possibly swaps in the set of selected features due to high correlation between the covariates, which in this case can occur. Our reference model approach applied to the stepwise backward regression achieves outstanding improvements considering its simplicity, yet it does not reach the goodness of more thoughtful implementations, as projpred (see results of Section \ref{projection}).

@@

This paper has discussed the importance of using a reference model when dealing with feature selection, or more in general model reduction. We want to highlight that the goal of this paper is not to provide a specific algorithm or procedure to carry out the selection, but rather to bring motivations and attentions to the general use of a reference model. 
\\
The reference model acts as an approximation of the data generation mechanism through its predictive distribution. Such approximation is generally less noisy than the sample estimation available from the observed data, leading to the main benefits of the reference model approach. In our comparisons, we have analysed the effect of a reference model in the form of a filter on the observed target values on top of different variable selection methods widely used. The overall benefits we have showed independently of the specific method applied, consisting in stability and goodness of the selection, motivate what we refer to as the reference model approach for the model reduction framework. Such approach translates into a family of different methods, some of which have been present in the literature for several years.
\\
In a real-world data application (body fat data), we have applied one of these methods, namely the projection predictive approach as indicated by \cite{paper:projpred}, in opposition to the selection via stepwise backward regression with the results provided by \cite{paper:bodyfat}. The selection via projection resulted to be more stable and led to a sparser representation. After adding additional noisy covariates, the selected submodel showed also higher predictive performance compared to the one selected by the stepwise backward regression, which included several non-relevant features. Applying a reference model on top of the stepwise backward regression resulted in a minor number of non-relevant features selected and an improved predictive performance, yet without reaching projection predictive approach results.
\\
Whenever it is possible to come up with a reasonable reference model, our suggestion is to employ it when selecting models, because of its nature as baseline of the comparisons and its ability to improve the selection. Note that one of the main challenges in many real world application will consist in devising the reference model itself and assessing its predictive performance, for which we suggest the user to rely on robust modelling workflow indications \cite[see e.g.][]{gabry2019visualization}.
\\
All Bayesian models have been implemented using Stan \citep{paper:stan}, which uses HMC-NUTS sampler \citep{hoffman2014no} and allows full posterior Bayesian inference. The code to run all the experiments is available at \url{https://github.com/fpavone/ref-approach-paper}.

@@

\hypertarget{ap:shrinkage_signal}{%
\section{Appendix: Shrinkage factors with Dirichlet-Laplace prior}\label{ap:shrinkage_signal}}
Figure \ref{fig:k_DL} shows the posterior mean shrinkage factors using the Dirichlet-Laplace prior for the normal means problem in the simulated data study of Section \ref{comparison}. The result is qualitatively equal to the regularised horseshoe prior's one (see Figure \ref{fig:k}).
